Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-08-04 resubmission/01-introduction.qmd
update resubmit

The justification of economic inequality has a long tradition in sociology, with seminal works such as the ones from Kluegel [-@kluegel_social_1995a; -@kluegel_beliefs_1981c] and Lane [@lane_political_1962; @lane_market_1986]. Lane's main aim was to "help explain the tenacious hold of market capitalism on the public mind" (@lane_market_1986, p.384), proposing the distinction between market and political justice. Whereas political justice refer to government actions aligned with justice principles such equality or need, market justice involves the allocation linked to earned desserts, particularly through individual contributions based on as effort and ability. Sociological survey research in this area has considered questions related to the extent to which it is considered just to get better access to social services (such as pensions, health and education) based on payment capacity. Although such services could be considered basic citizens rights, the evidence suggest that a relevant share of the population justify inequality in these domains [@lindh_public_2015]. In the present paper, we argue that there are two relevant factors to take into account when analyzing differences between individuals and groups to this regard: the perception of economic inequality, and the perception of meritocracy.

The perception of economic inequality can be understood as an individual's subjective assessment of how resources are allocated among members of a given society [@akyelken_urban_2020]. Regardless of their measurement, various studies have shown that this perception often underestimates the gap between the rich and the poor, revealing inconsistencies between objective and subjective aspects of inequality [@trump_income_2018]. Moreover, there is consistent evidence that perceived inequality, even under control of objective inequality, could have implications for attitudes toward the distribution of goods and services [@castillo_perception_2022; @schroder_income_2017]. redistributive preferences, justification of inequality, and legitimacy of the economic system [@castillo_perception_2022; @garcia-sanchez_perceptions_2018; @garcia-sanchez_attitudes_2020]. From this perspective, in this paper we argue that perceptions of economic inequality may play a relevant aspect when understanding market justice preferences, as a lower perceived inequality could reinforce the idea that the differences in access to social services could be justified --- a link that becomes difficult to hold when high inequality is perceived.

Along with perception of inequality, a second focus of this paper is the perception of meritocracy. Meritocracy posits that inequality can be legitimized through distributive criteria such as effort and talent [@young_rise_1962; @davis_principles_2001]. Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality as these beliefs are associated with attitudes that legitimize status differences [@batruch_belief_2023]. In a context as Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@boccardo_30_2020], meritocratic perceptions have the potential to affect the acceptance of social inequalities. Therefore, incorporating the perception of meritocracy in this literature allows for an understanding of how attitudes toward inequality are shaped not only by the perception of economic gaps, but also by beliefs about how these gaps are justified. Furthermore, these beliefs are often consolidated from an early age, reinforced by institutions promoting values such as effort and individual skills to climb socially [@castillo_socialization_2024; @reynolds_perceptions_2014].

At this point we can propose the following research questions: To what extent the preferences for market justice are associated with perceptions of inequality and meritocracy? How stable are these associations (within subjects) along time? The primary objective of this study is to analyze the interplay between perceptions of inequality and meritocracy and their joint influence on preferences for market justice along time in Chile using the panel survey ELSOC (Social Chilean Longitudinal Study, 2016-2023). The chilean case appears as a particular case for studying market justice preferences given their story of high inequality and strong free-market policies introduced in the dictatorship period [@boccardo_30_2020; @flores_top_2020]. Additionally, this analysis seeks to explore how political and social contingencies---such as the 2019 and 2022 social movements---might have moderated these relationships by prompting more critical reflection on the commodification of social services. Recognizing the temporal dimension in shaping market justice preferences is essential, given that historical and contextual factors that could challenge or reaffirm the way in which individuals persceive inequality and meritocracy, and consequantely --- following our argument --- the preferences for market justice.

------------------------------------------------------------------------

Papelera

Research shows that greater perceived inequality fosters stronger preferences for redistribution, regardless of actual inequality levels [@garcia-sanchez_vicious_2019; @gimpelson_misperceiving_2018]. However, other findings suggest that heightened perceptions of income gaps between high- and low-paying occupations may lead to greater justification of inequality [@Castillo2011].

A less explored area in this regard concerns market justice, which denotes the degree to which individuals consider just that social goods and services (e.g., health, pensions, education) are allocated according to individual contribution, competition, and ability to pay [@lane_market_1986; @lindh_public_2015; @kluegel_legitimation_1999]. Market justice attitudes reflect the belief that the market promotes procedural fairness---equality of opportunity---so that outcomes result from individual achievement [@lane_market_1986]. Indeed, Lindh \[-\] shows that market justice preferences are stronger in countries with higher private spending on services than in those with more comprehensive welfare systems.

[lo de la interacción, me parece que es too much para la introducción]
The perception of meritocracy and the perception of economic inequality would interact intricately in shaping preferences for market justice. On the one hand, a stronger perception of merit as a criterion for distribution could minimize the perception of economic gaps, thus justifying an unequal system [@castillo_percepcion_2019; @mijs_paradox_2021]. On the other hand, the perception of inequality could moderate the impact of meritocracy, as a higher perception of economic gaps might challenge the idea that these are solely based on merit.

>>>>

The justification of economic inequality has a long tradition in sociology, with seminal works such as the ones from Kluegel [-@kluegel_social_1995a; -@kluegel_beliefs_1986] and Lane [-@lane_political_1962; -@lane_market_1986]. Lane's main aim was to "help explain the tenacious hold of market capitalism on the public mind" [-@lane_market_1986, p.384], proposing the distinction between market and political justice. Whereas political justice refer to government actions aligned with justice principles such equality or need, market justice involves the allocation linked to earned desserts, particularly through individual contributions based on as effort and ability. Sociological survey research in this area has considered questions related to the extent to which it is considered just to get better access to social services (such as pensions, health and education) based on payment capacity [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015]. Although these services can be considered fundamental rights of citizenship, existing evidence indicates that a substantial share of the population legitimizes inequality in these domains, and that such preferences are shaped by individuals' socioeconomic characteristics as well as the broader contextual features of their countries [@koos_moral_2019; @lindh_public_2015; @busemeyer_skills_2014]. In the present paper, we argue that there are two other relevant factors to take into account when analyzing differences between individuals and groups to this regard: the perception of economic inequality and the perception of meritocracy.

The perception of economic inequality can be understood as an individual's subjective assessment of how resources are allocated among members of a given society [@akyelken_urban_2020]. Regardless of their measurement, various studies have shown that this perception often underestimates the gap between the rich and the poor, revealing inconsistencies between objective and subjective aspects of inequality [@trump_income_2018]. Moreover, there is consistent evidence that perceived inequality, even under control of objective inequality, could have implications for attitudes toward the distribution of goods and services [@castillo_perception_2022; @schroder_income_2017], redistributive preferences, justification of inequality, and legitimacy of the economic system [@castillo_perception_2022; @garcia-sanchez_perceptions_2018; @garcia-sanchez_attitudes_2020]. From this perspective, in this paper we argue that perceptions of economic inequality may play a relevant aspect for understanding market justice preferences, as a lower perceived inequality could reinforce the idea that the differences in access to social services could be justified --- a link that becomes difficult to hold when high inequality is perceived.

Along with perception of inequality, a second focus of this study is the perception of meritocracy. Meritocracy posits that inequality can be legitimized through distributive criteria such as effort and talent [@young_rise_1962; @davis_principles_2001]. Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality as these beliefs are associated with attitudes that legitimize status differences [@batruch_belief_2023]. In a context as Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@madariaga_three_2020; @boccardo_30_2020], meritocratic perceptions have the potential to affect the acceptance of social inequalities. Therefore, incorporating the perception of meritocracy in this literature allows for an understanding of how attitudes toward inequality are shaped not only by the perception of economic gaps, but also by beliefs about how these gaps are justified. Furthermore, these beliefs are often consolidated from an early age, reinforced by institutions promoting values such as effort and individual skills to climb socially [@castillo_socialization_2024; @reynolds_perceptions_2014].

At this point we can propose the following research questions: To what extent the preferences for market justice are associated with perceptions of inequality and meritocracy? How stable are these associations (within subjects) along time? The primary objective of this study is to examine the interplay between perceptions of inequality and beliefs in meritocracy, and how their interaction shapes preferences for market justice over time in Chile, using data from the ELSOC panel survey (Chilean Longitudinal Social Study, 2016–2023). The chilean context appears as a particular case for studying market justice preferences given their story of high inequality and strong free-market policies introduced in the dictatorship period and expanded under democratic governments [@madariaga_three_2020; @flores_top_2020]. Additionally, this analysis seeks to explore how political and social contingencies---such as the 2019 and 2022 social movements---might have moderated these relationships by prompting more critical reflection on the commodification of social services [@somma_no_2021]. Recognizing the temporal dimension in shaping market justice preferences is essential, given that historical and contextual factors that could challenge or reaffirm the way in which individuals persceive inequality and meritocracy, and consequantely --- following our argument --- the preferences for market justice.

### 2025-08-04 resubmission/02-antecedents.qmd
update resubmit

Conceptually, *market justice* has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit. It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who makes a contrast between market justice and political justice. The author defines market justice as a system of "earned deserts", whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills. In contrast, political justice emphasizes principles of equality and need, which are often represented by the welfare state action through social policies. An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement [@lane_market_1986]. Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open. In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of individual responsibility.

@@

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution. This is particularly salient in countries with more restricted public provision of social services. For example, Castillo et al. [-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that right-wing individuals are more supportive of market-based distribution of welfare. Beyond individual characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [@campbell_policy_2012; @pierson_when_1993].

@@

Perceptions of inequality have been associated with attitudes about market justice [@lindh_public_2015; @kluegel_social_1995a]. Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].ƒ

Perceptions of inequality refer to individuals' subjective evaluations of the extent, causes, and consequences of income and wealth disparities. Unlike objective measures such as the Gini index, perceived inequality captures how individuals make sense of distributive hierarchies in their everyday lives, shaped by reference groups, social comparisons, and information environments [@mijs_stratified_2016; @gimpelson_misperceiving_2018; @garcia-castro_perceiving_2020]. Scholars have proposed multiple dimensions of perceived inequality, including its magnitude (how significant are the gaps), vertical structure (between which groups), the trend over time (increasing or decreasing), and legitimacy (whether it is just or not) [@engelhardt_what_2018; @garcia-sanchez_vicious_2019]. These dimensions encompass both cognitive and normative aspects of perceptions of inequality and can vary across societies and social groups, depending on exposure, ideology, and personal experience [@castillo_perception_2022; @garcia-sanchez_perceptions_2018].

@@

In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups. Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023]. This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds. Moreover, @busemeyer_positive_2021 argues that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures. This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.

## The case of Chile

Chile offers an interesting case for examining how public attitudes toward the distribution of social services evolve amid declining poverty and persistently high income inequality within a residual welfare framework [@huber_democracy_2012]. Following the neoliberal reforms of the 1980s, the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@ferre_welfare_2023]. This model has favored the affluent, while lower-income groups are left dependent on limited and underfunded public alternatives. Research on social stratification characterizes Chile as a society where mobility into intermediate classes is feasible, but access to elite strata remains largely restricted [@torche_unequal_2005]. While some indicators—such as education and occupation—suggest a relatively fluid class structure, income mobility remains constrained [@EspinozaNunez2014].
Despite steady economic expansion, Chile continues to rank among the most unequal nations in the OECD, marked by a high Gini coefficient and substantial wealth concentration at the top [@flores_top_2019]. In this context, Chile experienced a wave of mass protests from October 2019 to March 2020. Initially led by high school students, the demonstrations quickly gained momentum as broader segments of society mobilized around demands for more equality of opportunities, particularly regarding precarious access to welfare services. The political system interpreted this unrest as a call to reform existing political institutions, with a special focus on equalizing opportunities and providing more egalitarian access to welfare by reducing reliance on out-of-pocket mechanisms [@somma_no_2021].

@@

In sum, Chile represents an illustrative case of a particular form of *moral economy* [@koos_moral_2019], in which national socioeconomic characteristics, institutional arrangements, and public attitudes coexist. This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system. At the same time, perceptions of inequality are high, and there is widespread public criticism of the commodification of social services. For the above reasons, we argue that the Chilean case represents a relevant context to study the role of perceived income inequality and meritocracy in shaping public preferences regarding the commodification of welfare services, particularly from the perspective of individual change.

@@

Building upon the previous literature, this study proposes that attitudes toward market justice are shaped by a dynamic interplay between individuals' perceptions of economic inequality and their beliefs in meritocracy. Specifically, we argue that both perceptions independently and interactively influence the extent to which individuals endorse market-based distributions of social goods and services in Chile.

First, consistent with previous findings, we expect that a higher perception of economic inequality will be associated with lower market justice preferences. When individuals perceive smaller income gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing the acceptance of outcomes based on competition and ability to pay. Conversely, a heightened perception of inequality may erode confidence in market fairness, weakening support for market-based distribution. This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force in shaping public attitudes toward inequality and justice.

Second, higher perceived meritocracy is expected to be positively associated with market justice preferences. Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit. This aligns with the idea that meritocratic perceptions serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit. This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.

@@

Additionally, this study examines whether these relationships vary over time, particularly in light of broader social and political transformations. We are especially interested in whether changes in individuals’ perceptions of inequality and meritocracy are associated with shifts in their market justice attitudes across different time points. The primary focus of this study is thus to scrutinize how individual-level changes in perceptions relate to changes in preferences over time. While we acknowledge the relevance of contextual events — such as the 2019 political outburst or the subsequent constitutional process — for public opinion and political orientation [@disi_pavlic_justification_2025], identifying causal effects of such events (e.g., mass protest exposure) on individual attitudes lies outside the scope of this research. Instead, we focus on how intra-individual trajectories in perceived inequality and merit shape evolving attitudes toward the commodification of social services.

@@

$H1$: Higher perceived economic inequality is associated with less market justice preferences.

$H2$: Higher meritocratic perceptions are associated with higher market justice preferences.

$H3$: The positive association between meritocratic perceptions and market justice preferences is moderated by perceived economic inequality; specifically, this association is weaker when perceived economic inequality is high.

$H4a$: Changes in perceived income inequality are less strongly associated with changes in market justice attitudes over time.

$H4b$: Changes in meritocracy beliefs are less strongly associated with changes in market justice attitudes over time.

>>>>

Conceptually, *market justice* has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit. It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who makes a contrast between market justice and political justice. The author defines market justice as a system of "earned deserts", whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills. In contrast, political justice emphasizes principles of equality and need, which are often represented by the welfare state action through social policies. An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement and procedures [@lane_market_1986]. Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open. In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of individual responsibility [@kluegel_legitimation_1999; @lindh_public_2015].

@@

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019; @lindh_public_2015]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution. This is particularly salient in countries with more restricted public provision of social services. For example, Castillo et al. [-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that right-wing individuals are more supportive of market-based distribution of welfare. Beyond individual characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [@campbell_policy_2012; @pierson_when_1993].

@@

Perceptions of inequality have been associated with a range of distributive attitudes [@castillo_perception_2022; @garcia-sanchez_vicious_2019; @trump_income_2018; @mccall_exposure_2017]. Research indicates that perceptions of inequality influence redistributive preferences to a greater extent than objective or actual economic inequality [@gimpelson_misperceiving_2018]. Others have shown that the perception of inequality, through the perceived wage gap, is related to the justification of greater inequality [@Castillo2011]. In that sense, lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@garcia-sanchez_vicious_2019].

Perceptions of inequality refer to individuals' subjective evaluations of the extent, causes, and consequences of income and wealth disparities [@akyelken_urban_2020]. Unlike objective measures such as the Gini index, perceived inequality captures how individuals make sense of distributive hierarchies in their everyday lives, shaped by reference groups, social comparisons, and information environments [@mijs_stratified_2016; @gimpelson_misperceiving_2018; @garcia-castro_perceiving_2020]. Scholars have proposed multiple dimensions of perceived inequality, including its magnitude (how significant are the gaps), vertical structure (between which groups), the trend over time (increasing or decreasing), and legitimacy (whether it is just or not) [@engelhardt_what_2018; @garcia-sanchez_vicious_2019]. These dimensions encompass both cognitive and normative aspects of perceptions of inequality and can vary across societies and social groups, depending on exposure, ideology, and personal experience [@castillo_perception_2022; @garcia-sanchez_perceptions_2018].

@@

In addition to influencing individual attitudes toward inequality, meritocratic perceptions can contribute to social division and the stigmatization of disadvantaged groups. Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023]. This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds. Moreover, @busemeyer_positive_2021 argues that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures. This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.

## The Chilean context

Chile offers an interesting case for examining how public attitudes toward the distribution of social services evolve amid declining poverty and persistently high income inequality within a residual welfare framework [@ferre_welfare_2023]. Following the neoliberal reforms of the 1980s introduced during the civil-military dictatorship (1973-1989), the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@madariaga_three_2020; @boccardo_30_2020]. This model has favored the affluent, while lower-income groups are left dependent on limited and underfunded public alternatives. Research on social stratification characterizes Chile as a society where mobility into intermediate classes is feasible, but access to elite strata remains largely restricted [@lopez-roldan_comparative_2021; @torche_intergenerational_2014a]. While some indicators—such as education and occupation—suggest a relatively fluid class structure, income mobility remains constrained [@espinoza_movilidad_2014].

Despite steady economic expansion, Chile continues to rank among the most unequal nations in the OECD, marked by a high Gini coefficient and substantial wealth concentration at the top [@flores_top_2020]. In this context, Chile experienced a wave of mass protests from October 2019 to March 2020. Initially led by high school students, the demonstrations quickly gained momentum as broader segments of society mobilized around demands for more equality of opportunities, particularly regarding precarious access to welfare services. The political system interpreted this unrest as a call to reform existing political institutions, with a special focus on equalizing opportunities and providing more egalitarian access to welfare by reducing reliance on out-of-pocket mechanisms [@somma_no_2021].

@@

In sum, Chile represents an illustrative case of a particular form of *moral economy* [@koos_moral_2019], in which national socioeconomic characteristics, institutional arrangements, and public attitudes coexist. This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system [@madariaga_three_2020]. At the same time, perceptions of inequality are high, and there is widespread public criticism of the commodification of social services [@somma_no_2021]. For the above reasons, we argue that the Chilean case represents a relevant context to study the role of perceived income inequality and meritocracy in shaping public preferences regarding the commodification of welfare services, particularly from the perspective of individual change.

@@

Building upon the previous literature, this study proposes that attitudes toward market justice are shaped by a dynamic interplay between individuals' perceptions of economic inequality and their perceptions of meritocracy. Specifically, we argue that both perceptions independently and interactively influence the extent to which individuals endorse market-based distributions of social goods and services in Chile.

First, consistent with previous findings about redistributive preferences [@castillo_perception_2022], we expect that a higher perception of economic inequality will be associated with lower market justice preferences. When individuals perceive smaller income gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing the acceptance of outcomes based on competition and ability to pay. Conversely, a heightened perception of inequality may erode confidence in market fairness, weakening support for market-based distribution. This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force in shaping public attitudes toward inequality and justice [@canalesceron_sujeto_2021; @mac-clure_justicia_2024].

Second, higher perceived meritocracy is expected to be positively associated with market justice preferences. Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit [@castillo_socialization_2024]. This aligns with the idea that meritocratic perceptions serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit. This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.

@@

Additionally, this study examines whether these relationships vary over time, particularly in light of broader social and political transformations. We are especially interested in whether changes in individuals’ perceptions of inequality and meritocracy are associated with shifts in their market justice attitudes across different time points. The primary focus of this study is thus to scrutinize how individual-level changes in perceptions relate to changes in preferences over time. While we acknowledge the relevance of contextual events — such as the 2019 political outburst or the subsequent constitutional process — for public opinion and political orientation [@disipavlic_justification_2025], identifying causal effects of such events (e.g., mass protest exposure) on individual attitudes lies outside the scope of this research. Instead, we focus on how intra-individual trajectories in perceived inequality and merit shape evolving attitudes toward the commodification of social services.

@@

$H1_{a} \; (\text{Between-person}):$ Individuals who perceive higher levels of economic inequality will report lower support for market justice preferences, on average, compared to those who perceive lower levels of inequality.

$H1_{b} \; (\text{Within-person}):$ Increases in an individual's perception of economic inequality over time will be associated with decreases in their support for market justice preferences.

$H2_{a} \; (\text{Between-person}):$ Individuals who endorse stronger meritocratic perceptions will report higher support for market justice preferences, on average, compared to those with weaker meritocratic perceptions

$H2_{b} \; (\text{Within-person}):$ Increases in an individual's meritocratic perceptions over time will be associated with increases in their support for market justice preferences.

$H3_{a} \; (\text{Between-person}):$ The positive association between meritocratic perceptions and market justice preferences will be weaker among individuals who perceive higher levels of economic inequality, compared to those who perceive lower inequality.

$H3_{b} \; (\text{Within-person}):$ The positive association between meritocratic perceptions and market justice preferences will be weaker at times when individuals perceive higher levels of inequality, relative to times when they perceive lower inequality.

$H4_{a} \; (\text{Growth moderation – Inequality}):$ The longitudinal association between perceived economic inequality and market justice preferences will strengthen over time, both within and between individuals.

$H4_{b} \; (\text{Growth moderation – Meritocracy}):$ The longitudinal association between meritocratic perceptions and market justice preferences will weaken over time, both within and between individuals.

### 2025-08-04 resubmission/paper-resubmit.qmd
update resubmit

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "../input/bib/harvard.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra-Sanhueza
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  This study investigates the relationship between perceptions of economic inequality, meritocratic perceptions, and preferences for market justice in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. Results indicate that rising perceptions of inequality are associated with lower support for market justice; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness. These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
  \newline
  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---

>>>>

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "../input/bib/harvard.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra-Sanhueza
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
 Several countries have experienced a shift toward the privatization and commodification of public goods, welfare policies, and social services. In Latin America, Chile stands out as a paradigmatic case where this trend has led to the extensive marketization of essential services. From a moral economy perspective, the extent to which individuals consider it fair for access to such services to depend on market criteria has been conceptualized as market justice preferences. This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and market justice preferences in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey – ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. The findings show that greater perceived inequality is associated with lower market justice preferences. However, individuals who believe that effort is rewarded are more likely to legitimize existing disparities. In contrast, the perception that talent is rewarded shows a negative effect on market justice preferences; an effect that intensifies as perceived inequality increases over time. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse on justice and fairness. These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks
  \newline
 **Keywords**: Inequality perception, meritocracy, commodification, Chile, public preferences 
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---
